Supplemental Materials: \\ \emph{Active contours-driven registration method for the structure-informed segmentation of diffusion MR images}

### 2015-10-01 Publications/2015-NeuroImage/04-methods.tex
fix labels with prefix

% -*- root: 00-main.tex -*-
\section{Methods}\label{sec:methods}

\subsection{Registration framework and segmentation model}\label{sec:methods_map}
Let $\gammaset_R = \{\Gamma_m: m \in \mathbb{N}, m \leq N_S\}$ be the set of $N_S$ surfaces
  extracted from the undistorted \gls*{t1} image (the reference space $R$).
We reformulate the segmentation of the distorted \gls*{dmri} images (the moving space $M$)
  as a registration problem where we search for an underlying deformation field $U$ such that
  the structures in $R$ defined by $\gammaset_R$ align optimally with their corresponding
  structures in $M$:

  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+\vec{u}(\vec{r}),
  \label{eq:transform}
  \end{align}
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ denotes the dimensionality of images.
Finally, $\vec{u} = \vec{u}(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
The general overview of how the surfaces interact with the registration framework
  is presented in \autoref{fig:method}.

\begin{figure}
  \includegraphics[width=\linewidth]{figure01}
  \caption{The active contours are defined as the interfacing surfaces of the competing
    \glspl{roi} $\Omega_l$, which
  are represented in green and dark blue in this close-up.
  They evolve iteratively following their inward normals $\hat{n}_i$ at each vertex
    $\vec{v}_i$ of the mesh.
  The gradient speeds $\bar{s}_i$ drive registration, which are computed as the disparity of the data
    energies with respect to the two limiting regions of $M(\vec{v}_i)$, the features of the image
    $M$ in the location of vertex $\vec{v}_i$ (see \suppl{equation SM5}).
  In this figure, $\bar{s}_1$ is written in the lower
    box, with $\Omega_{wm}$ in the inner limiting region, $\Omega_{gm}$ the outer region, and
    $w_{0,1}$ is the relative area associated with vertex $\vec{v}_1$ with respect to
    the total area of surface $\Gamma_0$.
      }\label{fig:method}
\end{figure}

\paragraph*{Cost-function derivation}
In a Bayesian framework for registration \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014},
  the mappings $U$ in \eqref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $M$.
Let $\omegaset = \{\Omega_l: l \in \mathbb{N}, l \leq N_L\}$ be the set of $N_L$ competing regions in
  which $M$ is partitioned by the projection of $\gammaset_R$.
Using Bayes' rule, the posterior likelihood is computed as:

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:bayes_rule}
  \end{equation}
  where $P(M \mid U,\omegaset)$ is the data likelihood.
Because $\omegaset$ are mapped by $U$, we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.
The best estimate $\hat{U}$ then satisfies the maximum a posteriori criterion
  \citep{bishop_pattern_2006} and it aligns $\gammaset_R$ into $M$.
First, we assume independence between pixels, and thus we break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the moving image.
For convenience and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999a,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with the parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$.
\newcomment[RV\#1(C.4)]{%
Even though the features being segmented are not generally i.i.d., the spatial interdependency of
  voxels is implicitly supported by the piecewise smooth partition of the space \omegaset{}.
In \autoref{fig:model} it is shown that marginal distributions of data can be approximated by
  multivariate normal distributions.} Therefore,

  \begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) =   \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the image $M$.
\newcomment[RV\#1(C.2)]{%
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized. In \autoref{fig:model},
  this minimization is illustrated for the segmentation of the \gls*{fa} and the \gls*{adc} maps
  of one subject.}

\begin{figure}
\includestandalone[width=\linewidth]{figure02-model}
% \includegraphics[width=\linewidth]{figure04A}
% \includegraphics[width=\linewidth]{figure04B}
% \includegraphics[width=\linewidth]{figure04C}
\caption{%
\newcomment{%
Evolution of the segmentation model defined by the homogeneous regions $\Omega_l$
  for a real dataset.
A) shows the initialization of the model, when the anatomically-correct surfaces are not
  mapped into the \gls*{dmri} space.
The panel shows the resulting conditional distributions of segmenting the (distorted) \gls*{fa} and
  \gls*{adc} maps with the original (undistorted) surfaces.
Panel B shows the equivalent information, using the surfaces with perfect matching with the distorted space
  of \gls*{dmri}.
Since the surfaces are now aligned in $M$, the spreads of the conditional distributions for each region are
  reduced with respect to those in A.
Panel C, after \regseg{} registration, shows how the covariance matrices $\boldsymbol{\Sigma}_l$
  are minimized with respect to panel A, and the resulting segmentation is close
  to the theoretical situation presented in panel B.
The left (large) plot in each panel represents a kernel density estimate of the location and spread of
  each region $\Omega_l$ in the bivariate feature space of $M$, which comprises
  the \gls*{fa} and \gls*{adc} maps.
At the top and the right margins, the marginal distributions of $\Omega_l$ are also plotted for
  the \gls*{fa} and \gls*{adc}, respectively.
On the right, the sampled distribution of each region is shown in small scatter plot, in contrast to
  the complete bivariate data.}
}\label{fig:model}
\todo[inline]{RV\#1(C.5)}
\end{figure}

@@

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \text{ with} \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:thikonov}
  \end{align}
 which requires that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Finally, the maximum a posteriori problem is adapted to a variational problem where we search for
  the minimum of an energy functional by applying $E(\vec{u}) = -\log \{P( M \mid U) \, P(U)\}$:

  \begin{align}
  E(\vec{u}) &= -\log \underset{l}{\prod}
  \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) = \notag\\ &=
  \const + \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, +   \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}.
  \label{eq:energy}
  \end{align}
This expression is the dual of the Mumford-Shah functional that corresponds
  to the framework of \acrlong*{acwe} \citep{chan_active_2001}
  with the anisotropic regularization term of \cite{nagel_investigation_1986}.

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}

\paragraph*{Deformation model}\label{sec:deformation_model}
Since the vertices of the surfaces $\{\vec{v}_i: \vec{v}_i \subset \gammaset \}_{i=1 \ldots N_V}$
  are probably located off-grid, it is necessary to derive $\vec{u}_i = \vec{u}(\vec{v}_i)$ from a discrete set of parameters
  $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved using a set of associated basis functions $\psi_k$ \eqref{eq:nodes_tfm}.
In our implementation, $\psi_k$ is a tensor-product B-spline kernel of degree three.

  \begin{equation}
  \vec{v}_i' = \vec{v}_i + \vec{u}_i = \vec{v}_i + \sum_k \psi_k(\vec{r}) \: \vec{u}_k.
  \label{eq:nodes_tfm}
  \end{equation}

\paragraph*{Optimization}
\label{sec:gradient_descent}
To find the minimum of the energy functional \eqref{eq:energy},
  we propose a gradient-descent approach with respect to the underlying
  deformation field using the following \gls*{pde}:

  \begin{equation}
  \frac{\partial \vec{u}(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
  where $t$ is an artificial time parameter of the contour
  evolution and $\vec{u}_k$ are the parameters that support the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that the anisotropy is aligned with the imaging axes to simplify
  \eqref{eq:energy} as expression \eqref{eq:app_energy} in \ref{app:reg_term},
  and thus to compute its derivative \eqref{eq:general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, +   \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\},
  \label{eq:gradient_descent}
  \end{align}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The computation of the explicit shape gradients at each $\vec{v}_i'$ is illustrated in \autoref{fig:method}.
Then, \eqref{eq:gradient_descent} can be reformulated as (see \suppl{equations SM6-SM12}):

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \vec{g}_k  + \boldsymbol{\alpha} \cdot \vec{u}_k - \boldsymbol{\beta} \cdot (\Delta \vec{u}_k).
  \label{eq:final_gradient}
  \end{align}

@@

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t - \vec{g}_k\} }
                    {\mathcal{F}\{(\delta^{-1} + \boldsymbol{\alpha})\, I - \boldsymbol{\beta}\Delta\}} \right\},
  \label{eq:update_equation}
  \end{align}
  where $I$ denotes the identity operator.

\paragraph*{Implementation details and convergence}
\label{sec:conv_report}
The \regseg{} tool includes a multiresolution strategy on the free-form deformation field.
Registration pyramids are created by setting the spacing between the control points of the B-spline basis
  functions for each level of the multiresolution strategy.
The implementation details as well as other features (such as the sparse matrix approach
  to fast interpolation) and the main parameters
  (such as $\delta$, $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, the B-spline grid resolutions,
  and target image smoothing) are discussed in \suppl{section S1}.

\subsection{Evaluation protocol}\label{sec:evaluation_protocol}
In order to assess the performance of \regseg{}, we defined the following general
  evaluation protocol:
1) Extract the set of undistorted surfaces $\gammaset_R$;
2) Compute a ground-truth field of displacements $U_\text{true}$, which is applied to
  generate warped images ($M$) for segmentation;
3) Execute \regseg{} with $\gammaset_R$ and use the warped data as inputs; and
4) Perform a visual assessment and compute the error metrics.

@@

\subsection{Simulated phantoms}\label{sec:digital_phantoms}
The workflow required to simulate the digital phantoms and to assess the performance of
  \regseg{} with them is presented in \autoref{fig:evphantoms}.
A set of four binary objects (i.e., ``box'', ``ball'', ``L'',
  and ``gyrus'') was generated by combining the binarization of
  analytical shapes and mathematical morphology.
The reference surfaces $\gammaset_R$ were extracted from the binary shapes
  using \emph{FreeSurfer} tools \citep{fischl_freesurfer_2012}.
The ground-truth distortion was generated using a chain of two displacement
  fields supported by grids of B-spline basis functions.
The coefficients of the basis functions were generated randomly for
  both levels in their three dimensions.
The three components of the displacements $\vec{u} = (u_d)$
  were bounded above by 40\% of the separation between the control points
  at each level to obtain diffeomorphic transforms
  after concatenation \citep{rueckert_diffeomorphic_2006}.
The first deformation field was applied to generate large warpings
  with control points separated by 50.50 mm in the three dimensions
  ($u_d\leq$ 20.20 mm).
With the second warping, we aimed to obtain a field with smoothness
  close to that found in a typical distortion field of \gls*{dmri} data
  \citep{irfanoglu_susceptibility_2011}.
Therefore, the control points were separated by 25.25 mm ($u_d\leq$ 10.10 mm).
After generating the ground-truth deformation, the original surfaces
  were warped by interpolating the displacements field at each vertex.
The warped surfaces $\gammaset_\text{true}$ were binarized to generate tissue fractions
  at low (\isores[mm]{2.0}) and high (\isores[mm]{1.0}) resolutions.
Using a \gls*{mr} simulator \citep{caruyer_phantomas_2014}, we synthesized
  \gls*{t1} (TE/TR= 10/1500 ms) and \gls*{t2} images (TE/TR= 90/5000 ms), which
  corresponded to each phantom type, with each at two resolutions
  (1.0 mm and 2.0 mm isotropic).
The field of view at both resolutions was \isores[mm]{100}.
Next, \regseg{} was applied to map $\gammaset_R$ onto the warped phantoms to
  obtain the registered surfaces ($\hat{\gammaset}_\text{test}$).
To quantify the misregistration error, we computed the Hausdorff distance between
 $\hat{\gammaset}_\text{test}$ and $\gammaset_\text{true}$ using \citep{commandeur_vtk_2011}.
In total, 1200 experiments (four phantom types $\times$ 150 random warpings $\times$ two resolutions) were
  performed according to the workflow illustrated in \autoref{fig:evphantoms}.

\begin{figure*}
\includegraphics[width=\linewidth]{figure02}
\caption{Evaluation of \regseg{} using phantom data according to the following instrumental workflow.
  1) The reference surfaces $\gammaset_R$ are triangularized meshes extracted from the four binary shapes (i.e., ``box'', ``ball'', ``L'', ``gyrus'').
  2) A ground-truth displacement field was generated as described in \autoref{sec:digital_phantoms}, and applied to warp
      $\gammaset_R$, thereby obtaining $\gammaset_\text{true}$.
  3) After being warped, $\gammaset_\text{true}$ were projected onto the corresponding discrete 3D volume and downsampled to create partial volume effects at two resolutions,
     i.e., \isores[mm]{2.0} and \isores[mm]{1.0}, thereby producing sets of tissue fractions maps.
  4) The tissue fractions were fed into a \acrfull*{mr} simulator, which generated \acrfull*{t1} and \acrfull*{t2} -like images at the
     two possible resolutions.
  5) The \regseg{} tool was applied using the warped test images as multispectral moving images and $\gammaset_R$ as shape priors.
  6) The agreement between the surfaces fitted by \regseg{} ($\hat{\gammaset}_\text{test}$) and $\gammaset_\text{true}$ were assessed
     visually using automatically generated visual reports and quantitativelly with the Hausdorff distance between the
     corresponding surfaces.}\label{fig:evphantoms}
\end{figure*}

>>>>

% -*- root: 00-main.tex -*-
\section{Methods}\label{sec:regseg-methods}

\subsection{Registration framework and segmentation model}\label{sec:regseg-methods_map}
Let $\gammaset_R = \{\Gamma_m: m \in \mathbb{N}, m \leq N_S\}$ be the set of $N_S$ surfaces
  extracted from the undistorted \gls*{t1} image (the reference space $R$).
We reformulate the segmentation of the distorted \gls*{dmri} images (the moving space $M$)
  as a registration problem where we search for an underlying deformation field $U$ such that
  the structures in $R$ defined by $\gammaset_R$ align optimally with their corresponding
  structures in $M$:

  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+\vec{u}(\vec{r}),
  \label{eq:regseg-transform}
  \end{align}
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ denotes the dimensionality of images.
Finally, $\vec{u} = \vec{u}(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
The general overview of how the surfaces interact with the registration framework
  is presented in \autoref{fig:regseg-method}.

\begin{figure}
  \includegraphics[width=\linewidth]{figure01}
  \caption{The active contours are defined as the interfacing surfaces of the competing
    \glspl{roi} $\Omega_l$, which
  are represented in green and dark blue in this close-up.
  They evolve iteratively following their inward normals $\hat{n}_i$ at each vertex
    $\vec{v}_i$ of the mesh.
  The gradient speeds $\bar{s}_i$ drive registration, which are computed as the disparity of the data
    energies with respect to the two limiting regions of $M(\vec{v}_i)$, the features of the image
    $M$ in the location of vertex $\vec{v}_i$ (see \suppl{equation SM5}).
  In this figure, $\bar{s}_1$ is written in the lower
    box, with $\Omega_{wm}$ in the inner limiting region, $\Omega_{gm}$ the outer region, and
    $w_{0,1}$ is the relative area associated with vertex $\vec{v}_1$ with respect to
    the total area of surface $\Gamma_0$.
      }\label{fig:regseg-method}
\end{figure}

\paragraph*{Cost-function derivation}
In a Bayesian framework for registration \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014},
  the mappings $U$ in \eqref{eq:regseg-transform} are
  evaluated based on their posterior probability given the observed data
  $M$.
Let $\omegaset = \{\Omega_l: l \in \mathbb{N}, l \leq N_L\}$ be the set of $N_L$ competing regions in
  which $M$ is partitioned by the projection of $\gammaset_R$.
Using Bayes' rule, the posterior likelihood is computed as:

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:regseg-bayes_rule}
  \end{equation}
  where $P(M \mid U,\omegaset)$ is the data likelihood.
Because $\omegaset$ are mapped by $U$, we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.
The best estimate $\hat{U}$ then satisfies the maximum a posteriori criterion
  \citep{bishop_pattern_2006} and it aligns $\gammaset_R$ into $M$.
First, we assume independence between pixels, and thus we break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:regseg-bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:regseg-transform} in the moving image.
For convenience and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999a,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with the parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$.
\newcomment[RV\#1(C.4)]{%
Even though the features being segmented are not generally i.i.d., the spatial interdependency of
  voxels is implicitly supported by the piecewise smooth partition of the space \omegaset{}.
In \autoref{fig:regseg-model} it is shown that marginal distributions of data can be approximated by
  multivariate normal distributions.} Therefore,

  \begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) =   \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:regseg-pdf}
  \end{align}
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the image $M$.
\newcomment[RV\#1(C.2)]{%
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized. In \autoref{fig:regseg-model},
  this minimization is illustrated for the segmentation of the \gls*{fa} and the \gls*{adc} maps
  of one subject.}

\begin{figure}
\includestandalone[width=\linewidth]{figure02-model}
% \includegraphics[width=\linewidth]{figure04A}
% \includegraphics[width=\linewidth]{figure04B}
% \includegraphics[width=\linewidth]{figure04C}
\caption{%
\newcomment{%
Evolution of the segmentation model defined by the homogeneous regions $\Omega_l$
  for a real dataset.
A) shows the initialization of the model, when the anatomically-correct surfaces are not
  mapped into the \gls*{dmri} space.
The panel shows the resulting conditional distributions of segmenting the (distorted) \gls*{fa} and
  \gls*{adc} maps with the original (undistorted) surfaces.
Panel B shows the equivalent information, using the surfaces with perfect matching with the distorted space
  of \gls*{dmri}.
Since the surfaces are now aligned in $M$, the spreads of the conditional distributions for each region are
  reduced with respect to those in A.
Panel C, after \regseg{} registration, shows how the covariance matrices $\boldsymbol{\Sigma}_l$
  are minimized with respect to panel A, and the resulting segmentation is close
  to the theoretical situation presented in panel B.
The left (large) plot in each panel represents a kernel density estimate of the location and spread of
  each region $\Omega_l$ in the bivariate feature space of $M$, which comprises
  the \gls*{fa} and \gls*{adc} maps.
At the top and the right margins, the marginal distributions of $\Omega_l$ are also plotted for
  the \gls*{fa} and \gls*{adc}, respectively.
On the right, the sampled distribution of each region is shown in small scatter plot, in contrast to
  the complete bivariate data.}
}\label{fig:regseg-model}
\todo[inline]{RV\#1(C.5)}
\end{figure}

@@

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \text{ with} \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:regseg-thikonov}
  \end{align}
 which requires that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Finally, the maximum a posteriori problem is adapted to a variational problem where we search for
  the minimum of an energy functional by applying $E(\vec{u}) = -\log \{P( M \mid U) \, P(U)\}$:

  \begin{align}
  E(\vec{u}) &= -\log \underset{l}{\prod}
  \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) = \notag\\ &=
  \const + \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, +   \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}.
  \label{eq:regseg-energy}
  \end{align}
This expression is the dual of the Mumford-Shah functional that corresponds
  to the framework of \acrlong*{acwe} \citep{chan_active_2001}
  with the anisotropic regularization term of \cite{nagel_investigation_1986}.

\subsection{Numerical Implementation}
\label{sec:regseg-numerical_implementation}

\paragraph*{Deformation model}\label{sec:regseg-deformation_model}
Since the vertices of the surfaces $\{\vec{v}_i: \vec{v}_i \subset \gammaset \}_{i=1 \ldots N_V}$
  are probably located off-grid, it is necessary to derive $\vec{u}_i = \vec{u}(\vec{v}_i)$ from a discrete set of parameters
  $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved using a set of associated basis functions $\psi_k$ \eqref{eq:regseg-nodes_tfm}.
In our implementation, $\psi_k$ is a tensor-product B-spline kernel of degree three.

  \begin{equation}
  \vec{v}_i' = \vec{v}_i + \vec{u}_i = \vec{v}_i + \sum_k \psi_k(\vec{r}) \: \vec{u}_k.
  \label{eq:regseg-nodes_tfm}
  \end{equation}

\paragraph*{Optimization}
\label{sec:regseg-gradient_descent}
To find the minimum of the energy functional \eqref{eq:regseg-energy},
  we propose a gradient-descent approach with respect to the underlying
  deformation field using the following \gls*{pde}:

  \begin{equation}
  \frac{\partial \vec{u}(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:regseg-general_gradient_descent}
  \end{equation}
  where $t$ is an artificial time parameter of the contour
  evolution and $\vec{u}_k$ are the parameters that support the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that the anisotropy is aligned with the imaging axes to simplify
  \eqref{eq:regseg-energy} as expression \eqref{eq:regseg-app_energy} in \ref{app:reg_term},
  and thus to compute its derivative \eqref{eq:regseg-general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, +   \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\},
  \label{eq:regseg-gradient_descent}
  \end{align}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The computation of the explicit shape gradients at each $\vec{v}_i'$ is illustrated in \autoref{fig:regseg-method}.
Then, \eqref{eq:regseg-gradient_descent} can be reformulated as (see \suppl{equations SM6-SM12}):

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \vec{g}_k  + \boldsymbol{\alpha} \cdot \vec{u}_k - \boldsymbol{\beta} \cdot (\Delta \vec{u}_k).
  \label{eq:regseg-final_gradient}
  \end{align}

@@

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t - \vec{g}_k\} }
                    {\mathcal{F}\{(\delta^{-1} + \boldsymbol{\alpha})\, I - \boldsymbol{\beta}\Delta\}} \right\},
  \label{eq:regseg-update_equation}
  \end{align}
  where $I$ denotes the identity operator.

\paragraph*{Implementation details and convergence}
\label{sec:regseg-conv_report}
The \regseg{} tool includes a multiresolution strategy on the free-form deformation field.
Registration pyramids are created by setting the spacing between the control points of the B-spline basis
  functions for each level of the multiresolution strategy.
The implementation details as well as other features (such as the sparse matrix approach
  to fast interpolation) and the main parameters
  (such as $\delta$, $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, the B-spline grid resolutions,
  and target image smoothing) are discussed in \suppl{section S1}.

\subsection{Evaluation protocol}\label{sec:regseg-evaluation_protocol}
In order to assess the performance of \regseg{}, we defined the following general
  evaluation protocol:
1) Extract the set of undistorted surfaces $\gammaset_R$;
2) Compute a ground-truth field of displacements $U_\text{true}$, which is applied to
  generate warped images ($M$) for segmentation;
3) Execute \regseg{} with $\gammaset_R$ and use the warped data as inputs; and
4) Perform a visual assessment and compute the error metrics.

@@

\subsection{Simulated phantoms}\label{sec:regseg-digital_phantoms}
The workflow required to simulate the digital phantoms and to assess the performance of
  \regseg{} with them is presented in \autoref{fig:regseg-evphantoms}.
A set of four binary objects (i.e., ``box'', ``ball'', ``L'',
  and ``gyrus'') was generated by combining the binarization of
  analytical shapes and mathematical morphology.
The reference surfaces $\gammaset_R$ were extracted from the binary shapes
  using \emph{FreeSurfer} tools \citep{fischl_freesurfer_2012}.
The ground-truth distortion was generated using a chain of two displacement
  fields supported by grids of B-spline basis functions.
The coefficients of the basis functions were generated randomly for
  both levels in their three dimensions.
The three components of the displacements $\vec{u} = (u_d)$
  were bounded above by 40\% of the separation between the control points
  at each level to obtain diffeomorphic transforms
  after concatenation \citep{rueckert_diffeomorphic_2006}.
The first deformation field was applied to generate large warpings
  with control points separated by 50.50 mm in the three dimensions
  ($u_d\leq$ 20.20 mm).
With the second warping, we aimed to obtain a field with smoothness
  close to that found in a typical distortion field of \gls*{dmri} data
  \citep{irfanoglu_susceptibility_2011}.
Therefore, the control points were separated by 25.25 mm ($u_d\leq$ 10.10 mm).
After generating the ground-truth deformation, the original surfaces
  were warped by interpolating the displacements field at each vertex.
The warped surfaces $\gammaset_\text{true}$ were binarized to generate tissue fractions
  at low (\isores[mm]{2.0}) and high (\isores[mm]{1.0}) resolutions.
Using a \gls*{mr} simulator \citep{caruyer_phantomas_2014}, we synthesized
  \gls*{t1} (TE/TR= 10/1500 ms) and \gls*{t2} images (TE/TR= 90/5000 ms), which
  corresponded to each phantom type, with each at two resolutions
  (1.0 mm and 2.0 mm isotropic).
The field of view at both resolutions was \isores[mm]{100}.
Next, \regseg{} was applied to map $\gammaset_R$ onto the warped phantoms to
  obtain the registered surfaces ($\hat{\gammaset}_\text{test}$).
To quantify the misregistration error, we computed the Hausdorff distance between
 $\hat{\gammaset}_\text{test}$ and $\gammaset_\text{true}$ using \citep{commandeur_vtk_2011}.
In total, 1200 experiments (four phantom types $\times$ 150 random warpings $\times$ two resolutions) were
  performed according to the workflow illustrated in \autoref{fig:regseg-evphantoms}.

\begin{figure*}
\includegraphics[width=\linewidth]{figure02}
\caption{Evaluation of \regseg{} using phantom data according to the following instrumental workflow.
  1) The reference surfaces $\gammaset_R$ are triangularized meshes extracted from the four binary shapes (i.e., ``box'', ``ball'', ``L'', ``gyrus'').
  2) A ground-truth displacement field was generated as described in \autoref{sec:regseg-digital_phantoms}, and applied to warp
      $\gammaset_R$, thereby obtaining $\gammaset_\text{true}$.
  3) After being warped, $\gammaset_\text{true}$ were projected onto the corresponding discrete 3D volume and downsampled to create partial volume effects at two resolutions,
     i.e., \isores[mm]{2.0} and \isores[mm]{1.0}, thereby producing sets of tissue fractions maps.
  4) The tissue fractions were fed into a \acrfull*{mr} simulator, which generated \acrfull*{t1} and \acrfull*{t2} -like images at the
     two possible resolutions.
  5) The \regseg{} tool was applied using the warped test images as multispectral moving images and $\gammaset_R$ as shape priors.
  6) The agreement between the surfaces fitted by \regseg{} ($\hat{\gammaset}_\text{test}$) and $\gammaset_\text{true}$ were assessed
     visually using automatically generated visual reports and quantitativelly with the Hausdorff distance between the
     corresponding surfaces.}\label{fig:regseg-evphantoms}
\end{figure*}
